On the propriety of restricted Boltzmann machines

### 2016-11-20 writing/appendix/appendix-center.Rmd
more edits

---
title: "Supplement to: On the propriety of restricted Boltzmann machines"
author: "Andee Kaplan, Daniel Nordman, and Stephen Vardeman"
date: "November, 2016"
linestretch: 2 
output: 
  pdf_document:
    fig_caption: yes
    includes:
      in_header: ../latex/header.tex
    number_sections: yes
---

>>>>

---
title: "Supplementary Material to:  On the propriety of restricted Boltzmann machines"
author: "Andee Kaplan, Daniel Nordman, and Stephen Vardeman"
date: "November, 2016"
linestretch: 2 
output: 
  pdf_document:
    fig_caption: yes
    includes:
      in_header: ../latex/header.tex
    number_sections: yes
---

Supplement material to follow consists of Section \ref{on-the-one-to-one-correspondence-between-parameters-and-moments-in-the-rbm-distribution}, providing proofs and discussion of results relating to the data coding in restricted Boltzmann machine models (RBMs).

### 2016-11-20 writing/draft.Rmd
more edits

where $\boldsymbol \theta \equiv (\theta_{11}, \dots, \theta_{1\nh}, \dots, \theta_{\nv 1}, \dots, \theta_{\nv \nh}, \theta_{v_1}, \dots, \theta_{v_\nv}, \theta_{h_1}, \dots, \theta_{h_\nh}) \in \mathbb{R}^{\nv + \nh + \nv*\nh}$ denotes the vector of model parameters and the denominator $\gamma(\boldsymbol \theta) = \sum_{\boldsymbol x \in \mathcal{C}^{\nh+\nv}}\exp\left(\sum_{i = 1}^\nv \sum_{j=1}^\nh \theta_{ij} v_i h_j + \sum_{i = 1}^\nv\theta_{v_i} v_i + \sum_{j = 1}^\nh\theta_{h_j} h_j\right)$ is the normalizing function that ensures the probabilities (\ref{eqn:pmf}) sum to one. For 
\begin{align}
\boldsymbol x &= (h_1, \dots, h_\nh, v_1, \dots, v_\nh) \in \mathcal{C}^{\nv + \nh} \text{ and} \nonumber\\
\boldsymbol t(\boldsymbol x) &= (h_1, \dots, h_\nh, v_1, \dots, v_\nv, v_1 h_1, \dots, v_\nv h_\nh) \in \mathcal{C}^{\nh + \nv + \nh*\nv}, \label{eqn:t}
\end{align}
let $\mathcal{T} = \{\boldsymbol t(\boldsymbol x): \boldsymbol x \in \mathcal{C}^{\nv + \nh}\} \subset \mathbb{R}^{\nv + \nh + \nv * \nh}$ be the set of possible values for the vector of variables needed to compute probabilities (\ref{eqn:pmf}) in the model and write $Q_{\boldsymbol \theta}(\boldsymbol x) = \sum\limits_{i = 1}^\nh\sum\limits_{j=1}^\nv \theta_{ij} h_i v_j + \sum\limits_{i = 1}^\nh\theta_{hi} h_i + \sum\limits_{j = 1}^\nv\theta_{vj} v_j$ for the "neg-potential" function. The RBM model is parameterized by $\boldsymbol \theta$ containing two types of parameters, main effects and interaction effects. The main effects parameters ($\{\theta_{v_i}, \theta_{h_j}\}_{\substack{i = 1, \dots, \nv,\\j = 1, \dots, \nh}}$) weight the values of the visible $v_i$ and hidden $h_j$ nodes in probabilities (\ref{eqn:pmf}) and the interaction effects parameters ($\theta_{ij}$) weight the values of the connections $v_i h_j$, or dependencies, between hidden and visible layers.

@@

A RBM model can be thought to exhibit an analogous form of *near-degeneracy* when there is a disproportionate amount of probability placed on a small number of elements in the sample space of visibles and hiddens, $\mathcal{C}^{\nv + \nh}$. Using the idea of @handcock2003assessing, when the random vector $\boldsymbol t(\boldsymbol x) = (v_1, \dots, v_\nv, h_1, \dots, h_\nh, v_1 h_1, \dots, v_V h_\nh ) \in \mathcal{T} \equiv \{\boldsymbol t(\boldsymbol x): \boldsymbol x \in \mathcal{C}^{\nh + \nv} \}$ from (\ref{eqn:t}), appearing in the neg-potential function $Q_{\boldsymbol \theta}(\cdot)$, has a mean vector $\boldsymbol \mu(\boldsymbol \theta) \in \mathbb{R}^{\nv + \nh + \nv * \nh}$ close to the boundary of the convex hull of $\mathcal{T}$, then the RBM model can be said to exhibit near-degeneracy at $\boldsymbol \theta \in \mathbb{R}^{\nv + \nh + \nh*\nv}$. Here the mean of $\boldsymbol t(\boldsymbol x)$ is
$$
\boldsymbol \mu(\boldsymbol \theta) = \text{E}_{\boldsymbol \theta} \boldsymbol t(\boldsymbol X)  = \sum\limits_{x \in \mathcal{C}^{\nv + \nh}} \left\{ \boldsymbol t(\boldsymbol x) f_{\boldsymbol \theta} (\boldsymbol x) \right\} = \sum\limits_{x \in \mathcal{C}^{\nv + \nh}} \left\{ \boldsymbol t(\boldsymbol x)\dfrac{\exp\left(\sum\limits_{i = 1}^\nv \sum\limits_{j=1}^\nh \theta_{ij} v_i h_j + \sum\limits_{i = 1}^\nv\theta_{v_i} v_i + \sum\limits_{j = 1}^\nh\theta_{h_j} h_j\right)}{\sum\limits_{\boldsymbol x \in \mathcal{C}^{\nh + \nv}}\exp\left(\sum\limits_{i = 1}^\nv \sum\limits_{j=1}^\nh \theta_{ij} v_i h_j + \sum\limits_{i = 1}^\nv\theta_{v_i} v_i + \sum\limits_{j = 1}^\nh\theta_{h_j} h_j\right)}\right\}.
$$

@@

### Uninterpretability
For spatial Markov models, @kaiser2007statistical defines a measure of model impropriety he calls *uninterpretability*, which is characterized by dependence parameters in a model being so extreme that marginal mean-structures fail to hold as anticipated by a model statement. We adapt this notion to RBM models as follows. Note that in a RBM, the parameters $\theta_{v_1}, \dots, \theta_{v_\nv}$ and $\theta_{h_1}, \dots, \theta_{h_\nh}$ are naturally associated with main effects of visible and hidden variables and can be interpreted as (logit functions of) means for variables $V_1, \dots, V_\nv, H_1, \dots, H_\nh$ in a model with no interaction parameters, $\theta_{ij} = 0, i = 1, \dots, \nv, j = 1, \dots, \nh$. That is, with no interaction parameters, we have from (\ref{eqn:pmf}) that
$$
P_\theta(V_i=1) \propto e^{\theta_{v_i}} \quad \& \quad P_\theta(H_j=1) \propto e^{\theta_{h_j}}, \quad  i=1,\ldots,\nv,j=1,\ldots,\nh
$$
so that, for example, $logit(P_\theta(V_i=1)=\theta_{v_i}$ (or $2 \theta_{v_i}$) under the coding $\mathcal{C}=\{0,1\}$ (or $\{-1,1\}$). Hence, these main effect parameters have a clear interpretation under an independence model (one with $\theta_{ij} = 0$) but this interpretation can break down as interaction parameters increase in magnitude relative to the size of the main effects. In such cases, the main effect parameters $\theta_{v_1}$ and $\theta_{h_j}$ are no longer interpretable as originally intended in the models (statements of marginal means) and the dependence parameters are so large as to dominate the entire model probability structure (also destroying the model interpretation of dependence as local conditional modifications of an overall marginal mean structure). To assess which parameter values $\boldsymbol \theta$ may cause difficulties in interpretation, we use the difference $\text{E}\left[\boldsymbol X | \boldsymbol \theta\right] - \text{E}\left[\boldsymbol X | \boldsymbol \theta^* \right]$ between two model expectations: E$\left[\boldsymbol X | \boldsymbol \theta\right]$ at $\boldsymbol \theta$ and expectations E$\left[\boldsymbol X | \boldsymbol \theta^* \right]$ given independence where $\boldsymbol \theta^*$ matches $\boldsymbol \theta$ for all main effects but otherwise has $\theta_{ij} = 0$ for $i = 1, \dots, \nv, j = 1, \dots, \nh$. Hence, $\boldsymbol \theta^*$ and $\boldsymbol \theta$ have the same main effects but $\boldsymbol \theta^*$ has $0$ dependence parameters. Uninterpretability is then avoided at a parametric specification $\boldsymbol \theta$ if the model expected value at $\boldsymbol \theta$ is not very different from the corresponding model expectation under independence. Using this, it is possible to investigate what parametric conditions lead to uninterpretability in a model versus those that guarantee interpretable models. If $||\text{E}\left[\boldsymbol X | \boldsymbol \theta\right] - \text{E}\left[\boldsymbol X | \boldsymbol \theta^*\right]||$ is large, then the RBM model with parameter vector $\boldsymbol \theta$ is said to be uninterpetable. The quantities to compare in the RBM case are

@@

It should be noted that BwTNLV (2) and BwTNML (3) are drawing from the same stationary posterior distribution for vectors of visibles. The difference between these two methods is in how well the chains mix and how quickly they arrive at the target posterior distribution. After a burn-in period of 50 iterations selected by inspecting the trace plots, we assess the issue of mixing in two ways. First, the autocorrelation functions (ACF) for each possible vector of visibles are assessed and plotted in Figure \ref{fig:acf} with BwTNLV in black and BwTNML in red. As expected, ACF corresponding to the method that marginalizes out the hidden variables from the likelihood decreases to zero at a much faster rate, indicating better mixing for the chain.

@@

Secondly, we can assess the mixing of our chains using an idea of effective sample size. If the MCMC chain were truly iid draws from the target distribution, then for the parameter $p^{(i)}$ denoting the probability of the $i$th vector outcome for the four visibles $\mathbf{v}=(v_1,v_2,v_3,v_4)\in\{\pm 1\}^4$, $i=1,\ldots,16$ (i.e., computed from $\theta$ under (\ref{eqn:pmf})), its estimate as the average $\bar{p}^{(i)}$ of posterior sample versions would be distributed approximately Normal with mean given by the true posterior marginal probability of $p^{(i)}$, and variance $\sigma^2_i/M$, where $\sigma^2_i$ is the true posterior variance of $p^{(i)}$ and $M$ is the length of the chain. However, with the presence of correlation in our chain, the aysmptotic variance of $\bar{p}^{(i)}$ is instead approximately some $C_i/M$, where $C_i$ is some positive constant such that $C_i > \sigma^2_i$. We can use an overlapping block-means approach [@gelman2011inference] to get a crude estimate for $C_i$ as $\hat{C}_i = bS_b^2$, where $S_b^2$  denotes the sample variance of overlapping block means $\{\bar{p}_j^{(i)}=\sum_{k=j}^{j+b-1} p_k^{(i)}/b\}_{j=1}^{M-b+1}$ of length $b$ computed from the posterior samples $\{p_k^{(i)}\}_{k=1}^M$. We compare it to an estimate of $\sigma^2_i$ using sample variance $\hat{sigma}^2_i$ of the raw chain, $\{p_k^{(i)}\}_{k=1}^{M}$. Formally, we approximate the effective sample size of the length $M$ MCMC chain as

>>>>

where $\boldsymbol \theta \equiv (\theta_{11}, \dots, \theta_{1\nh}, \dots, \theta_{\nv 1}, \dots, \theta_{\nv \nh}, \theta_{v_1}, \dots, \theta_{v_\nv}, \theta_{h_1}, \dots, \theta_{h_\nh}) \in \mathbb{R}^{\nv + \nh + \nv*\nh}$ denotes the vector of model parameters and the denominator $\gamma(\boldsymbol \theta) = \sum_{\boldsymbol x \in \mathcal{C}^{\nh+\nv}}\exp\left(\sum_{i = 1}^\nv \sum_{j=1}^\nh \theta_{ij} v_i h_j + \sum_{i = 1}^\nv\theta_{v_i} v_i + \sum_{j = 1}^\nh\theta_{h_j} h_j\right)$ is the normalizing function that ensures the probabilities (\ref{eqn:pmf}) sum to one. For $\boldsymbol x = (h_1, \dots, h_\nh, v_1, \dots, v_\nh) \in \mathcal{C}^{\nv + \nh}$ and 
\begin{align}
\boldsymbol t(\boldsymbol x) &= (h_1, \dots, h_\nh, v_1, \dots, v_\nv, v_1 h_1, \dots, v_\nv h_\nh) \in \mathcal{C}^{\nh + \nv + \nh*\nv}, \label{eqn:t}
\end{align}
let $\mathcal{T} = \{\boldsymbol t(\boldsymbol x): \boldsymbol x \in \mathcal{C}^{\nv + \nh}\} \subset \mathbb{R}^{\nv + \nh + \nv * \nh}$ be the set of possible values for the vector of variables needed to compute probabilities (\ref{eqn:pmf}) in the model and write $Q_{\boldsymbol \theta}(\boldsymbol x) = \sum\limits_{i = 1}^\nh\sum\limits_{j=1}^\nv \theta_{ij} h_i v_j + \sum\limits_{i = 1}^\nh\theta_{hi} h_i + \sum\limits_{j = 1}^\nv\theta_{vj} v_j$ for the "neg-potential" function. The RBM model is parameterized by $\boldsymbol \theta$ containing two types of parameters, main effects and interaction effects. The main effects parameters ($\{\theta_{v_i}, \theta_{h_j}\}_{\substack{i = 1, \dots, \nv,\\j = 1, \dots, \nh}}$) weight the values of the visible $v_i$ and hidden $h_j$ nodes in probabilities (\ref{eqn:pmf}) and the interaction effects parameters ($\theta_{ij}$) weight the values of the connections $v_i h_j$, or dependencies, between hidden and visible layers.

@@

A RBM model can be thought to exhibit an analogous form of *near-degeneracy* when there is a disproportionate amount of probability placed on a small number of elements in the sample space of visibles and hiddens, $\mathcal{C}^{\nv + \nh}$. Using the idea of @handcock2003assessing, when the random vector $\boldsymbol t(\boldsymbol x) = (v_1, \dots, v_\nv, h_1, \dots, h_\nh, v_1 h_1, \dots, v_V h_\nh ) \in \mathcal{T} \equiv \{\boldsymbol t(\boldsymbol x): \boldsymbol x \in \mathcal{C}^{\nh + \nv} \}$ appearing in the neg-potential function $Q_{\boldsymbol \theta}(\cdot)$, has a mean vector $\boldsymbol \mu(\boldsymbol \theta) \in \mathbb{R}^{\nv + \nh + \nv * \nh}$ close to the boundary of the convex hull of $\mathcal{T}$, then the RBM model can be said to exhibit near-degeneracy at $\boldsymbol \theta \in \mathbb{R}^{\nv + \nh + \nh*\nv}$. Here the mean of $\boldsymbol t(\boldsymbol x)$ is
$$
\boldsymbol \mu(\boldsymbol \theta) = \text{E}_{\boldsymbol \theta} \boldsymbol t(\boldsymbol X)  = \sum\limits_{x \in \mathcal{C}^{\nv + \nh}} \left\{ \boldsymbol t(\boldsymbol x) f_{\boldsymbol \theta} (\boldsymbol x) \right\} = \sum\limits_{x \in \mathcal{C}^{\nv + \nh}} \left\{ \boldsymbol t(\boldsymbol x)\dfrac{\exp\left(\sum\limits_{i = 1}^\nv \sum\limits_{j=1}^\nh \theta_{ij} v_i h_j + \sum\limits_{i = 1}^\nv\theta_{v_i} v_i + \sum\limits_{j = 1}^\nh\theta_{h_j} h_j\right)}{\sum\limits_{\boldsymbol x \in \mathcal{C}^{\nh + \nv}}\exp\left(\sum\limits_{i = 1}^\nv \sum\limits_{j=1}^\nh \theta_{ij} v_i h_j + \sum\limits_{i = 1}^\nv\theta_{v_i} v_i + \sum\limits_{j = 1}^\nh\theta_{h_j} h_j\right)}\right\}.
$$

@@

### Uninterpretability
For spatial Markov models, @kaiser2007statistical defines a measure of model impropriety he calls *uninterpretability*, which is characterized by dependence parameters in a model being so extreme that marginal mean-structures fail to hold as anticipated by a model statement. We adapt this notion to RBM models as follows. Note that in a RBM, the parameters $\theta_{v_1}, \dots, \theta_{v_\nv}$ and $\theta_{h_1}, \dots, \theta_{h_\nh}$ are naturally associated with main effects of visible and hidden variables and can be interpreted as (logit functions of) means for variables $V_1, \dots, V_\nv, H_1, \dots, H_\nh$ in a model with no interaction parameters, $\theta_{ij} = 0, i = 1, \dots, \nv, j = 1, \dots, \nh$. That is, with no interaction parameters, we have from (\ref{eqn:pmf}) that
$$
P_\theta(V_i=1) \propto e^{\theta_{v_i}} \quad \text{ and } \quad P_\theta(H_j=1) \propto e^{\theta_{h_j}}, \quad  i=1,\ldots,\nv,j=1,\ldots,\nh
$$
so that, for example, $logit(P_\theta(V_i=1))=\theta_{v_i}$ (or $2 \theta_{v_i}$) under the coding $\mathcal{C}=\{0,1\}$ (or $\{-1,1\}$). Hence, these main effect parameters have a clear interpretation under an independence model (one with $\theta_{ij} = 0$) but this interpretation can break down as interaction parameters increase in magnitude relative to the size of the main effects. In such cases, the main effect parameters $\theta_{v_1}$ and $\theta_{h_j}$ are no longer interpretable as originally intended in the models (statements of marginal means) and the dependence parameters are so large as to dominate the entire model probability structure (also destroying the model interpretation of dependence as local conditional modifications of an overall marginal mean structure). To assess which parameter values $\boldsymbol \theta$ may cause difficulties in interpretation, we use the difference $\text{E}\left[\boldsymbol X | \boldsymbol \theta\right] - \text{E}\left[\boldsymbol X | \boldsymbol \theta^* \right]$ between two model expectations: E$\left[\boldsymbol X | \boldsymbol \theta\right]$ at $\boldsymbol \theta$ and expectations E$\left[\boldsymbol X | \boldsymbol \theta^* \right]$ given independence where $\boldsymbol \theta^*$ matches $\boldsymbol \theta$ for all main effects but otherwise has $\theta_{ij} = 0$ for $i = 1, \dots, \nv, j = 1, \dots, \nh$. Hence, $\boldsymbol \theta^*$ and $\boldsymbol \theta$ have the same main effects but $\boldsymbol \theta^*$ has $0$ dependence parameters. Uninterpretability is then avoided at a parametric specification $\boldsymbol \theta$ if the model expected value at $\boldsymbol \theta$ is not very different from the corresponding model expectation under independence. Using this, it is possible to investigate what parametric conditions lead to uninterpretability in a model versus those that guarantee interpretable models. If $||\text{E}\left[\boldsymbol X | \boldsymbol \theta\right] - \text{E}\left[\boldsymbol X | \boldsymbol \theta^*\right]||$ is large, then the RBM model with parameter vector $\boldsymbol \theta$ is said to be uninterpetable. The quantities to compare in the RBM case are

@@

It should be noted that BwTNLV (2.) and BwTNML (3.) are drawing from the same stationary posterior distribution for vectors of visibles. The difference between these two methods is in how well the chains mix and how quickly they arrive at the target posterior distribution. After a burn-in period of 50 iterations selected by inspecting the trace plots, we assess the issue of mixing in two ways. First, the autocorrelation functions (ACF) for each possible vector of visibles are assessed and plotted in Figure \ref{fig:acf} with BwTNLV in black and BwTNML in red. As expected, ACF corresponding to the method that marginalizes out the hidden variables from the likelihood decreases to zero at a much faster rate, indicating better mixing for the chain.

@@

Secondly, we can assess the mixing of our chains using an idea of effective sample size. If the MCMC chain were truly iid draws from the target distribution, then for the parameter $p^{(i)}$ denoting the probability of the $i$th vector outcome for the four visibles $\mathbf{v}=(v_1,v_2,v_3,v_4)\in\{\pm 1\}^4$, $i=1,\ldots,16$ (i.e., computed from $\boldsymbol \theta$ under (\ref{eqn:pmf})), its estimate as the average $\bar{p}^{(i)}$ of posterior sample versions would be distributed approximately Normal with mean given by the true posterior marginal probability of $p^{(i)}$, and variance $\sigma^2_i/M$, where $\sigma^2_i$ is the true posterior variance of $p^{(i)}$ and $M$ is the length of the chain. However, with the presence of correlation in our chain, the aysmptotic variance of $\bar{p}^{(i)}$ is instead approximately some $C_i/M$, where $C_i$ is some positive constant such that $C_i > \sigma^2_i$. We can use an overlapping block-means approach [@gelman2011inference] to get a crude estimate for $C_i$ as $\hat{C}_i = bS_b^2$, where $S_b^2$  denotes the sample variance of overlapping block means $\{\bar{p}_j^{(i)}=\sum_{k=j}^{j+b-1} p_k^{(i)}/b\}_{j=1}^{M-b+1}$ of length $b$ computed from the posterior samples $\{p_k^{(i)}\}_{k=1}^M$. We compare it to an estimate of $\sigma^2_i$ using sample variance $\hat{\sigma}^2_i$ of the raw chain, $\{p_k^{(i)}\}_{k=1}^{M}$. Formally, we approximate the effective sample size of the length $M$ MCMC chain as
